---
title: "ordination"
author: "Millie Chapman"
date: "7/8/2019"
output: html_document
---

```{r}
library(vegan)
library(tidyverse)
library(ggplot2)
```

```{r}
load("data/bryophytes.rda")

occ <- as.data.frame(bryophytes$Y)
traits <- as.data.frame(bryophytes$Tr)
phylo <- as.data.frame(bryophytes$Phylo)
```


```{r}
dist.bray<-vegdist(occ,binary="TRUE",method="bray")
dist.jacc<-vegdist(occ,binary="TRUE",method="jaccard") 
dist.euc<-vegdist(occ,binary="TRUE",method="euclidean") 
```


```{r}
# First step is to calculate a distance matrix. 
# Here we use Bray-Curtis distance metric
dist <- vegdist(occ,  method = "bray")

# PCoA is not included in vegan. 
library(ape)
PCOA <- pcoa(dist)

# plot the eigenvalues and interpret
barplot(PCOA$values$Relative_eig[1:10])
# Can you also calculate the cumulative explained variance of the first 3 axes?

# Some distance measures may result in negative eigenvalues--add a correction:
PCOA <- pcoa(dist, correction = "cailliez")

# Plot  results
biplot.pcoa(PCOA)
```



## Euclidean
  + Euclidean distance is a suitable multivariate measure of distance in situations where variables are expected to have equal variance. 
  + analyses using this metric are appropriate if all taxa are measured on comparable scales with comparable variability. 
  + On a mean–variance plot, the equal variance assumption corresponds to the assumption that the mean–variance relation is a horizontal line (slope zero).

## Bray-Curtis
  + standardisation used in the Bray–Curtis distance is only appropriate when between‐group differences in average absolute difference are matched by proportionate changes in average abundance. 
  + applies to averages across all taxa rather than to each taxon separately, because the summation occurs inside the fraction rather than outside it. 
  + Inside the fraction, there is no longer any consideration of the different variability of different taxa, as for the Euclidean and Manhattan distances. 
  + Bray–Curtis will only weight different taxa appropriately if all variables have similar variability. 
  + Bray-Curtis distance for abundance
    + When the distance metric is Euclidean, PCoA is equivalent to Principal Components Analysis. 
  
 ## Jaccard
  + Jaccard index for presence/absence data. 


